---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RAT

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

## Install as package
### 1. Create virtual environment with renv
Initialize a local R environment:
```{r, eval = FALSE}
# Open project in Rstudio
# Install and initialize renv 
install.packages("renv")
library(renv)
renv::init()
```

### 2. Install package from github
``` {r, eval = FALSE}
# To ensure Rstudio looks up BioConductor packages run:
setRepositories(ind = c(1:6, 8))
# Then install package with
devtools::install_github("XueningHe/RAT_package")
```


## Install as cloned repository

### 1. Clone repository
```{sh, eval = FALSE}
# In terminal at desired directory
git clone git@github.com:XueningHe/RAT_package.git
```

### 2. Install dependecies
```{r, eval = FALSE}
# Open RAT.Rproj in Rstudio

# To ensure Rstudio looks up BioConductor packages run:
setRepositories(ind = c(1:6, 8))
# Install renv
install.packages("renv")
library(renv)
# Restore from lock file
renv::restore()
```

### 3. Load package
```{r, eval = FALSE}
install.packages("devtools")
library(devtools)
devtools::load_all(getwd())
```

## Usage

### Load Shiny app

```{r, eval = FALSE}
library(RAT)

run_app()

```
*Some text about how the shiny app works.*


### Running without Shiny


**Load external data from directory "/RAT_package" for illustration.**
```{r}
# dmap will be the scaffold dataset

# load expression matrix
dmap_exprs <- read.csv("exprs_dmap.csv") # input data type can be anything like csv  
rownames(dmap_exprs) <- dmap_exprs[,1]   # rownames of expression matrix are gene names, colnames are sample names
dmap_exprs <- dmap_exprs[,-1]            # every element of matrix should be numbers.

# load phenotype table
dmap_pheno <- read.csv("pData_dmap.csv") # input data type can be anything like csv  
rownames(dmap_pheno) <- dmap_pheno[,1]   # rownames of phenotype table are sample names, colnames can be whatever.
dmap_pheno <- dmap_pheno[,-1,drop = F]   # drop=F prevents conversion from data.frame to vector

# check row-col concordance
# colnames of expression matrix should be identical to rownames of phenotype table
all(colnames(dmap_exprs)==rownames(dmap_pheno)) 

# ilaria will be new samples for projection

# load expression matrix
ilaria_exprs <- read.csv("exprs_ilaria.csv") # input data type can be anything like csv  
rownames(ilaria_exprs) <- ilaria_exprs[,1]   # rownames of expression matrix are gene names, colnames are sample names
ilaria_exprs <- ilaria_exprs[,-1]            # every element of matrix should be numbers.

# load phenotype table
ilaria_pheno <- read.csv("pData_ilaria.csv")  # input data type can be anything like csv  
rownames(ilaria_pheno) <- ilaria_pheno[,1]    # rownames of phenotype table are sample names, colnames can be whatever.
ilaria_pheno <- ilaria_pheno[,-1,drop = F]    # drop=F prevents conversion from data.frame to vector

# check row-col concordance
# colnames of expression matrix should be identical to rownames of phenotype table
all(colnames(ilaria_exprs)==rownames(ilaria_pheno)) 

```


Users mainly need 2 functions: `buildScaffold` and `projectSample`.


**`buildScaffold()`**

`buildScaffold` takes in an expression matrix (`exprs_scaffold`), a phenotype table (`pData_scaffold`), a column name within phenotype table (`group_scaffold`), and returns a scaffoldSpace object, containing a vector of names of differentially expressed genes in scaffold dataset, a gg object returned by ggplot, and a prcomp object returned by function \code{prcomp()}. Type `?buildScaffold` in console to see more optional parameters (which might be able to be adjusted in the webapp). Type `?scaffoldSpace` in console to see more information of the S4 class scaffoldSpace defined in this package.  
  
```{r}
library(RAT)
# To see the scaffold space 
space <- buildScaffold(dmap_exprs,dmap_pheno,"cell_types") # "cell_types" is a column name of dmap_pData

```
  
For now, calling buildScaffold will automatically print out the scaffold PCA plot. This can be changed in "R/buildScaffold.R".
  
**`projectSample`**
`projectSample()` takes in a scaffoldSpace object output by function `buildScaffold()`, an expression matrix, a corresponding phenotype table, and a column name. It outputs the scafffoled PCA plot, together with new samples

```{r}
projectSample(space,ilaria_exprs,ilaria_pheno,"cancer_type") # "cancer_type" is a column name of dmap_pData
```

