---
title: "Introduction to spaceRAT"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    highlight: pygments
author: "Søren Helweg Dam"
date: "Last updated: `r format(Sys.Date(), '%Y.%m.%d')`"
vignette: >
  %\VignetteIndexEntry{Introduction to spaceRAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

`spaceRAT` uses Ranked Analysis for Transcriptomes to project
any sample into a space of interest.
`spaceRAT` ranks the differentially expressed genes in your sample data
and uses that information to project it into a prebuilt or
user-built scaffold space. `spaceRAT` provides plotting functions that
allows the user to visually investigate their sample data.

In the following, a brief overview of the `spaceRAT` is illustrated. 
See the respective vignettes for additional information.


# Build scaffold

See `Build Scaffold` vignette for additional information. 
Here is a quick overview:

```{r tldr_scaffold}
library(spaceRAT)

# Load expression data
data("exprs_dmap", package="spaceRAT")
# Load phenotype data
data("pData_dmap", package="spaceRAT")
# Specify colname with phenotype information
colname <- "cell_types"

# Build scaffold
scaffold <- buildScaffold(
    exprs_dmap,
    data = "logged",
    pheno = pData_dmap,
    colname = colname
)

# Plot scaffold
plotScaffold(scaffold, title = "Scaffold plot")
```


# Project sample

See `Project Sample` vignette for additional information.
Here is a quick overview:

```{r tldr}
library(spaceRAT)

# Load count data
data("counts_ilaria", package="spaceRATScaffolds")

# Load custom or prebuilt scaffold
scaffold <- buildScaffold("DMAP_scaffold")

# Project sample
projectSample(
    space = scaffold,
    sample = counts_ilaria,
    title = "Samples projected into DMAP scaffold")
```

# Session Info

```{r session info}
sessionInfo()
```
