---
title: "Introduction to spaceRAT"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    highlight: pygments
author: "Søren Helweg Dam"
date: "Last updated: `r format(Sys.Date(), '%Y.%m.%d')`"
vignette: >
  %\VignetteIndexEntry{Introduction to spaceRAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

`spaceRAT` uses Ranked Analysis for Transcriptomes to project
any sample into a space of interest.
`spaceRAT` ranks the differentially expressed genes in your sample data
and uses that information to project it into a prebuilt or
user-built scaffold space. `spaceRAT` provides plotting functions that
allows the user to visually investigate their sample data.

In the following, a brief overview of the `spaceRAT` is illustrated. 
See the respective vignettes for additional information.

# Installation

*Dependencies*

```{r dependencies, eval = FALSE}
# Install Bioconductor dependencies
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("SummarizedExperiment", "S4Vectors"))

# Install spaceRATScaffolds from github
remotes::install_github("shdam/spaceRATScaffolds", build_vignettes = TRUE)
```


*Install from GitHub*
``` {r install, eval = FALSE}
remotes::install_github("XueningHe/spaceRAT", build_vignettes = TRUE)
```


# Build scaffold

See `Build Scaffold` vignette for additional information. 
Here is a quick overview:

```{r tldr_scaffold}
library(spaceRAT)

# Load expression data
data("DMAP_exprs", package="spaceRATScaffolds")
# Load phenotype data
data("DMAP_pData", package="spaceRATScaffolds")
# Specify colname with phenotype information
colname <- "cell_types"

# Build scaffold
scaffold <- buildScaffold(
    DMAP_exprs,
    data = "exprs",
    pheno = DMAP_pData,
    colname = colname
)

# Plot scaffold
plotScaffold(scaffold, title = "Scaffold plot")
```


# Project sample

See `Project Sample` vignette for additional information.
Here is a quick overview:

```{r tldr}
library(spaceRAT)

# Load count data
data("ilaria_counts", package="spaceRATScaffolds")

# Load custom or prebuilt scaffold
scaffold <- buildScaffold("DMAP_scaffold")

# Project sample
projectSample(
    space = scaffold,
    sample = ilaria_counts,
    title = "Samples projected into DMAP scaffold")
```

# Session Info

```{r session info}
sessionInfo()
```
