---
title: "Project Sample"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    highlight: pygments
author: "Søren Helweg Dam"
date: "Last updated: `r format(Sys.Date(), '%Y.%m.%d')`"
vignette: >
  %\VignetteIndexEntry{Project Sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Quick overview


```{r tldr}
library(spaceRAT)

# Load count data
data("counts_ilaria", package="spaceRATScaffolds")

# Load custom or prebuilt scaffold
scaffold <- buildScaffold("DMAP_scaffold")

# Project sample
projectSample(
    space = scaffold,
    sample = counts_ilaria,
    title = "Samples projected into DMAP scaffold")
```

See `?projectSample` for additional parameter settings.

# Introduction

In this vignette, we will demonstrate how to project a sample into a 
prebuilt scaffold space.


# Build or load scaffold

Prepare a scaffold space of interest. See the `Build scaffold` vignette for
information on how to build a custom scaffold.

For prebuilt scaffold, you can choose from the following scaffolds:

```{r list_scaffolds} 
library("spaceRATScaffolds")
listScaffolds()
```


Getting a prebuilt scaffold:

```{r setup}
library(spaceRAT)

scaffold <- buildScaffold("DMAP_scaffold")
```


# Load sample data


The sample data can have the data format `matrix`, `data.frame`, or 
`r  BiocStyle::Biocpkg("SummarizedExperiment")`.
It should have samples as columns and features as rows. Here is an example:

```{r ilexprs}
utils::data("counts_ilaria", "pData_ilaria",
            package = "spaceRATScaffolds")
counts_ilaria[1:5,1:5]
```

Optionally, you can add phenotype information to the projection. 
It could like this:

```{r ilpdata}
pData_ilaria[1:5,, drop=FALSE]
```
# Project sample

The projection is then performed by `projectSample`:

```{r project}
projectSample(
    space = scaffold, 
    sample = counts_ilaria,
    title = "Samples projected into DMAP scaffold")
```

With added phenotype information:
```{r projectwpheno}
projectSample(
    space = scaffold,
    sample = counts_ilaria,
    pheno = pData_ilaria,
    colname =  "cancer_type",
    title = "Samples projected into DMAP scaffold")

```

## Projecting a `r  BiocStyle::Biocpkg("SummarizedExperiment")`

If you have a `r  BiocStyle::Biocpkg("SummarizedExperiment")` object
with your sample data, the process is very similar.

```{r SE}
# Prepare data
se <- SummarizedExperiment::SummarizedExperiment(
    assays = S4Vectors::SimpleList("counts" = counts_ilaria),
    colData = S4Vectors::DataFrame(pData_ilaria)
)

# Build scaffold
projectSample(
    space = scaffold,
    sample = se,
    assay = "counts",
    colname =  "cancer_type",
    title = "Samples projected into DMAP scaffold")
```

See `?projectSample` for additional parameter settings.


# Session Info

```{r sessioninfo}
sessionInfo()
```
